\pkg{webchem}: An \proglang{R} Package to Retrieve Chemical Information from the Web

### 2019-10-07 revision3_2581/paper.tex

\documentclass[article, shortnames]{jss}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{Sweave}

\usepackage[utf8]{inputenc} %unicode support
\usepackage{booktabs}
\usepackage{fixltx2e}

@@

\Plainauthor{Eduard Sz\"ocs, Ralf B. Sch\"afer}
\Plaintitle{webchem: an R Package to Retrieve Chemical Information from the Web}
\Shorttitle{\pkg{webchem}: Chemical Information from the Web}

\Abstract{
A wide range of chemical information is freely available online, including identifiers, experimental and predicted chemical properties.
However, these data are scattered over various data sources and not easily accessible to researchers.
Manual searching and downloading of such data is time-consuming and error-prone.  
We developed the open-source \proglang{R} package \pkg{webchem} that allows users to automatically query chemical data from currently 11 web sources. 
These cover a broad spectrum of information.
The data are automatically imported into an \proglang{R} object and can directly be used in subsequent analyses.
\pkg{webchem} enables easy, structured and reproducible data retrieval and usage from publicly available web sources.
In addition, it facilitates data cleaning, identification and reporting of substances.
Consequently, it reduces the time researchers need to spend on chemical data compilation.
}
\Keywords{ecotoxicology, chemistry, data cleaning, web scraping, ropensci}
\Plainkeywords{ecotoxicology, chemistry, data cleaning, web scraping, ropensci}

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Eduard Sz\"ocs\\
  Institute for Environmental Sciences\\
  Universit\"at Koblenz-Landau\\
  Fortstraße 7\\
  76829 Landau, Germany\\
  E-mail: \email{szoecs@uni-landau.de}\\
  URL: \url{https://edild.github.io}
}

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

% citation alias for LUBW institution
\defcitealias{lubw_2016}{LUBW (2016)}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

When dealing with chemical data sets (e.g.\ environmental monitoring data, toxicological data), a first step is often to validate the names of chemicals or to link them to unique codes that simplify subsequent querying and appending of compound-related physico-chemical or toxicological information.
Several web sources provide chemical names or link them to unique codes (see also section \emph{Data sources} below).
However, manual searching for each compound, often through a graphical web interface, is tedious, error-prone and not reproducible \citep{Peng_2009}.

To simplify, robustify and automate this task, i.e.\ to search and retrieve chemical information from the web, we created the \pkg{webchem} package for the free and open source \proglang{R} language \citep{r_2015, Wehrens_2011}.
\proglang{R} is one of the most widely used software environments for data cleaning, analysing and visualising data, and supports full reproducibility of each step \citep{Marwick_2016}.

@@

\section[Implementation and design details]{Implementation and design details}
The \pkg{webchem} package is written entirely in \proglang{R} and available under a MIT license.
The development repository is hosted on \citet{github} and a stable version is released on the official \proglang{R} repository \citep{cran}.
\pkg{webchem} is part of the rOpenSci project \citep{boettiger2015building}, which aims at fully reproducible data analysis.

\pkg{webchem} follows best practices for scientific software \citep{wilson_best_2014, poisot_best_2015}, namely: (i) a public available repository with easy collaboration and an issue tracker (via GitHub), (ii) a non-restrictive license, version control (git), (iii) an elaborate test-suite covering more than 90\% of the relevant lines of code (currently approximately 1500 lines, using \pkg{testthat} \citep{wickham_testthat:_2011}), (iv) continuous integration (via \citet{travis-ci} and \citet{appveyor}; testing on Linux \& Windows with current and development \proglang{R} versions), (v) in-source documentation (using \pkg{roxygen2} \citep{wickham_roxygen2:_2015}) and (vi) compliance with a style guide \citep{wickham_advanced_2015}.

\pkg{webchem} builds on top of the following \proglang{R} packages:
\pkg{RCurl} \citep{lang_rcurl:_2015} and \pkg{httr} \citep{wickham_httr} for data transfer,
\pkg{stringr} \citep{wickham_stringr:_2015} for string handling,
\pkg{xml2} \citep{wickham_xml2} and \pkg{rvest} \citep{wickham_rvest} for parsing HTML and XML,
\pkg{jsonlite} \citep{ooms_jsonlite_2014} for parsing JSON,
\pkg{rcdk} \citep{guha_rcdk} for parsing SMILES.
For parsing molfiles we use a lightweight implementation of \citep{Grabner_Varmuza_Dehmer_2012}.

@@

Only few design decisions have been made:
Each function name has a prefix and suffix separated by an underscore \citep{Chamberlain_Szocs_2013}.
They follow the format of \code{source\_function}, e.g.\ \code{cs\_compinfo} uses ChemSpider as source (see next section) to retrieve compound information.
Some functions require querying first a unique identifier from the data source and then use this identifier to query further information.
The prefix \code{get} is used to denote these functions, e.g.\ \code{get\_csid} to retrieve the identifier used in ChemSpider.

@@

\section[Data sources]{Data sources}
The backbone of \pkg{webchem} are data sources providing their data and functionality to the public.
Currently, data can be retrieved from 11 sources.
These cover a broad spectrum of available data, like identifiers, experimental and predicted properties and regulatory information (Figure~\ref{fig:fig1}, a detailed overview of all sources is included as supplement):

\begin{description}
  \item[NIH Chemical Identifier Resolver (CIR) \citep{cir}]{A web service that converts from and to various chemical identifiers.}
  \item[Chemical Translation Service (CTS) \citep{Wohlgemuth_Haldiya_Willighagen_Kind_Fiehn_2010}]{A web service that converts from and to various chemical identifiers.}
  \item[ETOX \citep{etox}]{Information System Ecotoxicology and Environmental Quality Targets by the German Federal Environmental Agency. Provides basic identifiers, synonyms, ecotoxicological data and quality targets for different countries.}
  \item[PAN Pesticide Database \citep{pan}]{Information on pesticides - provides basic identifiers, ecotoxicological data and chemical properties.}
  \item[SRC Physprop \citep{physprop}]{Contains physical properties for over 41,000 chemicals.
  Physical properties collected from a wide variety of sources including experimental and modeled values.}
  \item[PubChem \citep{Kim_2016}]{PubChem is a public repository for information on chemical substances, providing identifiers, properties and synonyms.
  We use an interface to the PUG-REST web service \citep{Kim_Thiessen_Bolton_Bryant_2015}.}
  \item[Wikidata \citep{wiki}]{Wikipedia contains information for over 15,000 chemicals \citep{Ertl_Patiny_Sander_Rufener_Zasso_2015}. Currently \pkg{webchem} can only query chemical identifiers.}
  \item[Compendium of Pesticide Common Names \citep{wood}]{The compendium provides information on pesticide common names, identifiers and classification.}
  \item[ChemID\emph{plus} \citep{Tomasulo_2002}]{is a large web-based database provided by the National Library of Medicine. It provides identifiers, synonyms, toxicological data and chemical properties.}
  \item[ChemSpider \citep{pence_chemspider:_2010}]{is a free chemical structure database providing access to over 40 million structures. It provides identifiers, properties and can also be used to convert identifiers.}
  \item[OPSIN  \citep{Lowe_Corbett_Murray-Rust_Glen_2011}]{The Open Parser for Systematic IUPAC nomenclature is a chemical name interpreter and provides InChI and SMILES identifiers.}
\end{description}

@@

\begin{figure}[ht]
  \centering
  \includegraphics{fig1}
  \caption{Overview of current data sources. Input and output possibilities currently implemented in the package.}
  \label{fig:fig1}
\end{figure}

\section[Use cases]{Use cases}
\subsection[Install webchem]{Installation}
\pkg{webchem} can be easily installed and loaded from CRAN:

\begin{CodeChunk}
\begin{CodeInput}
R> install.packages("webchem")
R> library("webchem")
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\end{CodeChunk}

The package is under active development. The latest development version is available from GitHub and also permanently available at \citet{zenodo}.
This document has been created using \pkg{webchem} version 0.1.

@@

(i) \code{jagst}: This data set comprises environmental monitoring data of organic substances in the river Jagst, Germany, sampled in 2013.
The data is publicly available and can be retrieved from \citetalias{lubw_2016}.
It comprises concentrations  (in $\mathrm{\mu g~/~L}$) of  34 substances  on 13 sampling occasions.
First we load the data set and inspect the first six rows:

\begin{CodeChunk}
\begin{CodeInput}
R> data("jagst")
R> head(jagst)
\end{CodeInput}
\begin{CodeOutput}
        date          substance value qual
1 2013-01-04 2,4-Dimethylphenol 0.006    <
2 2013-01-29 2,4-Dimethylphenol 0.006    <
3 2013-02-26 2,4-Dimethylphenol 0.006    <
4 2013-03-26 2,4-Dimethylphenol 0.006    <
5 2013-04-23 2,4-Dimethylphenol 0.006    <
6 2013-05-22 2,4-Dimethylphenol 0.006    <
\end{CodeOutput}
\end{CodeChunk}

This data set identifies substances only by substance names. Values below the limit of quantification (LOQ) are indicated by a qualifier column.

(ii) \code{lc50}: This data consists of median acute lethal concentration for the water flea \textit{Daphnia magna} in 48 h tests ($LC_{50, D.magna, 48h}$) of 124 insecticides.
The data has been retrieved from the EPA ECOTOX database \citep{epa_2016}.

\begin{CodeChunk}
\begin{CodeInput}
R> data("lc50")
R> head(lc50)
\end{CodeInput}
\begin{CodeOutput}
       cas        value
4  50-29-3    12.415277
12 52-68-6     1.282980
15 55-38-9    12.168138
18 56-23-5 35000.000000
21 56-38-2     1.539119
36 57-74-9    98.400000
\end{CodeOutput}
\end{CodeChunk}

This data set identifies the substances only by CAS numbers.

@@

One possibility to resolve this, is to use different chemical identifiers allowing easy identification.
There are several identifiers available, e.g.\  registry numbers like CAS or EC, database identifiers like PubChemCID \citep{Kim_2016} or ChemSpiderID \citep{pence_chemspider:_2010}, line notations like SMILES \citep{Weininger_1990}, InChI and InChiKey \citep{Heller_McNaught_Pletnev_Stein_Tchekhovskoi_2015}. 
In this first example we query several identifiers to create a table that can be used as (i) supplemental information to a research article or (ii) to facilitate subsequent matching with other data.

@@

First we search for ETOX internal ID numbers using the substance names:

\begin{CodeChunk}
\begin{CodeInput}
R> subs <- unique(jagst$substance)
R> ids <- get_etoxid(subs, match = 'best')
R> head(ids)
\end{CodeInput}
\begin{CodeOutput}
  etoxid                           match distance                  query
1   8668     2,4-Dimethylphenol ( 8668 )        0     2,4-Dimethylphenol
2   8494 4-Chlor-2-methylphenol ( 8494 )        0 4-Chlor-2-methylphenol
3   <NA>                            <NA>     <NA>     4-para-nonylphenol
4   8397                Atrazin ( 8397 )        0                Atrazin
5   7240                 Benzol ( 7240 )        0                 Benzol
6   7331        Desethylatrazin ( 7331 )        0        Desethylatrazin
\end{CodeOutput}
\end{CodeChunk}

Only three substances could not be found in ETOX. 
Here we specify that only the \emph{`best'} match (in terms of the Levenshtein distance between query and results) is returned. 
A manual check confirms appropriate matches. 
Other options include: \emph{`all'} - returns all matches; \emph{`first'} - returns only the first match (not necessarily the best match); \emph{`ask'} - this enters an interactive mode, where the user is asked for a choice if multiple matches are found and \emph{`na'} which returns \code{NA} in case of multiple matches.

We use these data to retrieve basic information on the substances.

\begin{CodeChunk}
\begin{CodeInput}
R> etox_data <- etox_basic(ids$etoxid)
\end{CodeInput}
\end{CodeChunk}

\pkg{webchem} always returns a named list (one entry for each substance) and the available information content can be very voluminous.
Therefore, we provide extractor functions for the common identifiers: CAS, SMILES and InChIKeys.
\begin{CodeChunk}
\begin{CodeInput}
R> etox_cas <- cas(etox_data)
R> head(etox_cas)
\end{CodeInput}
\begin{CodeOutput}
       8668        8494        <NA>        8397        7240        7331 
 "105-67-9" "1570-64-5"          NA "1912-24-9"   "71-43-2" "6190-65-4" 
\end{CodeOutput}
\end{CodeChunk}

A variety of data are available and we cannot provide extractor functions for each of those.
Therefore, if users need to extract other data, they have to write simple extractor functions (see following examples).

In the same manner, we can now query other identifiers from another source using these CAS numbers (Figure~\ref{fig:fig1}), like PubChem
\begin{CodeChunk}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> cids <- get_cid(etox_cas)
R> pc_data <- pc_prop(cids, properties = c('CanonicalSMILES'))
R> pc_smiles <- smiles(pc_data)
\end{CodeInput}
\end{CodeChunk}

or ChemSpider
\begin{CodeChunk}
\begin{CodeInput}
R> csids <- get_csid(etox_cas, token = token)
R> cs_data <- cs_compinfo(csids, token = token)
R> cs_inchikey <- inchikey(cs_data)
\end{CodeInput}
\end{CodeChunk}

Finally, we combine the queried data into one data.frame
\begin{CodeChunk}
\begin{CodeInput}
R> res <- data.frame(name = subs, cas = etox_cas, smiles = pc_smiles, 
+    cid = pc_data$CID, inchikey = cs_inchikey, csid = cs_data$csid, 
+    stringsAsFactors = FALSE)
\end{CodeInput}
\end{CodeChunk}

Note that in order to use the ChemSpider functions, a personal authentication key (\code{token}) is needed, which can be retrieved from the ChemSpider web page. 
Finally, we obtain a compound table containing many different identifiers (Table~\ref{tab:comptable}), allowing easy identification and merging with other data sets, e.g.\ the \code{lc50} data set based on CAS.

\begin{table}[ht]
\centering
\begin{CodeChunk}
% latex table generated in R 3.3.1 by xtable 1.8-2 package
% Tue Sep 20 16:29:57 2016
\begin{tabular}{llllll}
  \toprule
Name & CAS & SMILES & CID & InChIKey & CSID \\ 
  \midrule
2,4-Dimethylphenol & 105-67-9 & CC1=CC(... & 7771 & KUFFULV... & 13839123 \\ 
  4-Chlor-2-methylphenol & 1570-64-5 & CC1=C(C... & 14855 & RHPUJHQ... & 14165 \\ 
  4-para-nonylphenol & - & - & - & - & - \\ 
  Atrazin & 1912-24-9 & CCNC1=N... & 2256 & MXWJVTO... & 2169 \\ 
  Benzol & 71-43-2 & C1=CC=C... & 241 & UHOVQNZ... & 236 \\ 
  Desethylatrazin & 6190-65-4 & CC(C)NC... & 22563 & DFWFIQK... & 21157 \\ 
   \bottomrule
\end{tabular}
\end{CodeChunk}
\caption{Identifiers for the jagst data sets as queried with \pkg{webchem}. Only the first 6 entries are shown. For SMILES and InChIKey only the first 7 characters are shown. - = not found.}
\label{tab:comptable}
\end{table}

\subsection[Toxicity of different pesticide groups]{Toxicity of different pesticide groups}
Another question we might ask is \emph{How does toxicity vary between insecticide groups?}
Answering this question would require tedious lookup of insecticide groups for each of the 124 CAS numbers in the \code{lc50} data set.
The Compendium of Pesticide Common Names \citep{wood} contains such information and can be easily queried using CAS numbers with \pkg{webchem}:

\begin{CodeChunk}
\begin{CodeInput}
R> aw_data <- aw_query(lc50$cas, type = 'cas')
\end{CodeInput}
\end{CodeChunk}

To extract the chemical group from the retrieved data set, we write a simple extractor function and apply this to the retrieved data:

\begin{CodeChunk}
\begin{CodeInput}
R> igroup <- sapply(aw_data, function(y) y$subactivity[1])
R> igroup[1:3]
\end{CodeInput}
\begin{CodeOutput}
                                  50-29-3 
            "organochlorine insecticides" 
                                  52-68-6 
               "phosphonate insecticides" 
                                  55-38-9 
"phenyl organothiophosphate insecticides" 
\end{CodeOutput}
\end{CodeChunk}

Figure \ref{fig:fig2} displays the result after additional data cleaning (see supplement for full code).
Overall, it took only 5 \proglang{R} statements to retrieve, clean and plot the data using \pkg{ggplot2} \citep{ggplot2}.

\begin{figure}[ht]
\begin{CodeChunk}

{\centering \includegraphics[width=0.7\textwidth]{plot_lc50-1}

}

\end{CodeChunk}
\caption{Toxicity of different pesticide groups. LC\textsubscript{50} values have been retrieved from EPA ECOTOX database, chemical groups from the Compendium of Pesticide Common Names \citep{wood}}
\label{fig:fig2}
\end{figure}

\subsection[Querying partitioning coefficients]{Querying partitioning coefficients}
Some data sources also provide data on chemical properties that can be queried.
Here we query for the \code{lc50} data the $\mathrm{log}~P_{oct/wat}$ from the SRC PHYSPROP database to build a simple quantitative structure–activity relationship (QSAR) to predict toxicity.

\begin{CodeChunk}
\begin{CodeInput}
R> pp_data <- pp_query(lc50$cas)
\end{CodeInput}
\end{CodeChunk}

>>>>

\documentclass[article]{jss}
\usepackage{thumbpdf,lmodern} 
\graphicspath{{Figures/}}

@@

\Plainauthor{Eduard Sz\"ocs, Ralf B. Sch\"afer}

\Plaintitle{webchem: An R Package to Retrieve Chemical Information from the Web}

\Shorttitle{\pkg{webchem}: Chemical Information from the Web}

\Abstract{ A wide range of chemical information is freely available
  online, including identifiers, experimental and predicted chemical
  properties.  However, these data are scattered over various data
  sources and not easily accessible to researchers.  Manual searching
  and downloading of such data is time-consuming and error-prone.  We
  developed the open-source \proglang{R} package \pkg{webchem} that
  allows users to automatically query chemical data from currently 11
  web sources.  These cover a broad spectrum of information.  The data
  are automatically imported into an \proglang{R} object and can
  directly be used in subsequent analyses.  \pkg{webchem} enables
  easy, structured and reproducible data retrieval and usage from
  publicly available web sources.  In addition, it facilitates data
  cleaning, identification and reporting of substances.  Consequently,
  it reduces the time researchers need to spend on chemical data
  compilation.  }

\Keywords{ecotoxicology, chemistry, data cleaning, web scraping, ropensci}

\Plainkeywords{ecotoxicology, chemistry, data cleaning, web scraping, ropensci}

%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

\Address{
  Eduard Sz\"ocs\\
  Institute for Environmental Sciences\\
  Universit\"at Koblenz-Landau\\
  Fortstra{\ss}e 7\\
  76829 Landau, Germany\\
  E-mail: \email{szoecs@uni-landau.de}\\
  URL: \url{https://edild.github.io}
}

% citation alias for LUBW institution
\defcitealias{lubw_2016}{LUBW (2016)}
% \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

When dealing with chemical data sets (e.g., environmental monitoring
data, toxicological data), a first step is often to validate the names
of chemicals or to link them to unique codes that simplify subsequent
querying and appending of compound-related physico-chemical or
toxicological information.  Several web sources provide chemical names
or link them to unique codes (see also
Section~\ref{sec:data-sources}).  However, manual searching for each
compound, often through a graphical web interface, is tedious,
error-prone and not reproducible \citep{Peng_2009}.

To simplify, robustify and automate this task, i.e., to search and
retrieve chemical information from the web, we created the
\pkg{webchem} package \citep{cran} for the free and open source
\proglang{R} language \citep{r_2015, Wehrens_2011}.  \proglang{R} is
one of the most widely used software environments for data cleaning,
analyzing and visualizing data, and supports full reproducibility of
each step \citep{Marwick_2016}.

@@

\section[Implementation and design details]{Implementation and design details}
The \pkg{webchem} package is written entirely in \proglang{R} and
available under an MIT license.  The development repository is hosted
on \citet{github} and a stable version is released on the
Comprehensive \proglang{R} Archive Network (CRAN) and available at
\url{https://CRAN.R-project.org/package=webchem}.  \pkg{webchem} is
part of the rOpenSci project \citep{boettiger2015building}, which aims
at fully reproducible data analysis.

\pkg{webchem} follows best practices for scientific software
\citep{wilson_best_2014, poisot_best_2015}, namely: (i) a public
available repository with easy collaboration and an issue tracker (via
GitHub), (ii) a non-restrictive license, version control (git), (iii)
an elaborate test-suite covering more than 90\% of the relevant lines
of code (currently approximately 1500 lines, using \pkg{testthat};
\citealt{wickham_testthat:_2011}), (iv) continuous integration (via
\citet{travis-ci} and \citet{appveyor}; testing on Linux \& Windows
with current and development \proglang{R} versions), (v) in-source
documentation (using \pkg{roxygen2}; \citealt{wickham_roxygen2:_2015})
and (vi) compliance with a style guide \citep{wickham_advanced_2015}.

\pkg{webchem} builds on top of the following \proglang{R} packages:
\pkg{RCurl} \citep{lang_rcurl:_2015} and \pkg{httr}
\citep{wickham_httr} for data transfer, \pkg{stringr}
\citep{wickham_stringr:_2015} for string handling, \pkg{xml2}
\citep{wickham_xml2} and \pkg{rvest} \citep{wickham_rvest} for parsing
HTML and XML, \pkg{jsonlite} \citep{ooms_jsonlite_2014} for parsing
JSON, \pkg{rcdk} \citep{guha_rcdk} for parsing SMILES.  For parsing
molfiles we use a lightweight implementation in package \pkg{RMol}
\citep{Grabner_Varmuza_Dehmer_2012}.

@@

Only few design decisions have been made: Each function name has a
prefix and suffix separated by an underscore
\citep{Chamberlain_Szocs_2013}.  They follow the format of
\code{source\_function}, e.g., \code{cs\_compinfo} uses ChemSpider as
source (see Section~\ref{sec:data-sources}) to retrieve compound
information.  Some functions require querying first a unique
identifier from the data source and then use this identifier to query
further information.  The prefix \code{get} is used to denote these
functions, e.g., \code{get\_csid} to retrieve the identifier used in
ChemSpider.

@@

\section[Data sources]{Data sources}\label{sec:data-sources}
The backbone of \pkg{webchem} are data sources providing their data
and functionality to the public.  Currently, data can be retrieved
from 11 sources.  These cover a broad spectrum of available data, like
identifiers, experimental and predicted properties and regulatory
information (a detailed overview of all sources is included in
Figure~\ref{fig:fig1}):

\begin{description}
\item[NIH Chemical Identifier Resolver (CIR)]{\citep{cir} A web
    service that converts from and to various chemical identifiers.}
\item[Chemical Translation Service (CTS)]{
  \citep{Wohlgemuth_Haldiya_Willighagen_Kind_Fiehn_2010} A web
    service that converts from and to various chemical identifiers.}
\item[ETOX]{\citep{etox} Information System Ecotoxicology and
    Environmental Quality Targets by the German Federal Environmental
    Agency. Provides basic identifiers, synonyms, ecotoxicological
    data and quality targets for different countries.}
\item[PAN Pesticide Database]{ \citep{pan} Information on pesticides --
    provides basic identifiers, ecotoxicological data and chemical
    properties.}
\item[SRC Physprop]{\citep{physprop} Contains physical properties for
    over 41,000 chemicals.  Physical properties collected from a wide
    variety of sources including experimental and modeled values.}
\item[PubChem]{\citep{Kim_2016} PubChem is a public repository for
    information on chemical substances, providing identifiers,
    properties and synonyms.  We use an interface to the PUG-REST web
    service \citep{Kim_Thiessen_Bolton_Bryant_2015}.}
\item[Wikidata]{\citep{wiki} Wikipedia contains information for over
    15,000 chemicals
    \citep{Ertl_Patiny_Sander_Rufener_Zasso_2015}. Currently
    \pkg{webchem} can only query chemical identifiers.}
\item[Compendium of Pesticide Common Names]{\citep{wood} The
    compendium provides information on pesticide common names,
    identifiers and classification.}
\item[ChemID\emph{plus}]{\citep{Tomasulo_2002} is a large web-based
    database provided by the National Library of Medicine. It provides
    identifiers, synonyms, toxicological data and chemical
    properties.}
\item[ChemSpider]{\citep{pence_chemspider:_2010} is a free chemical
    structure database providing access to over 40 million
    structures. It provides identifiers, properties and can also be
    used to convert identifiers.}
\item[OPSIN]{\citep{Lowe_Corbett_Murray-Rust_Glen_2011} The Open
    Parser for Systematic IUPAC nomenclature is a chemical name
    interpreter and provides InChI and SMILES identifiers.}
\end{description}

@@

\begin{figure}[t!]
  \centering
  \includegraphics{fig1}
  \caption{Overview of current data sources. Input and output possibilities currently implemented in the package.}
  \label{fig:fig1}
\end{figure}

\section[Use cases]{Use cases}
\subsection[Installation]{Installation}
\pkg{webchem} can be easily installed from CRAN and loaded:
%
\begin{CodeChunk}
\begin{CodeInput}
R> install.packages("webchem")
R> library("webchem")
\end{CodeInput}
\end{CodeChunk}
%
The package is under active development. The latest development
version is available from GitHub and also permanently available at
\citet{zenodo}.  This document has been created using \pkg{webchem}
version 0.4.0.

@@

(i) \code{jagst}: This data set comprises environmental monitoring data of organic substances in the river Jagst, Germany, sampled in 2013.
The data is publicly available and can be retrieved from \citetalias{lubw_2016}.
It comprises concentrations  (in $\mathrm{\mu g~/~L}$) of  34 substances  on 13 sampling occasions.
First we load the data set and inspect the first six rows:
%
\begin{CodeChunk}
\begin{CodeInput}
R> data("jagst", package = "webchem")
R> head(jagst)
\end{CodeInput}
\begin{CodeOutput}
        date          substance value qual
1 2013-01-04 2,4-Dimethylphenol 0.006    <
2 2013-01-29 2,4-Dimethylphenol 0.006    <
3 2013-02-26 2,4-Dimethylphenol 0.006    <
4 2013-03-26 2,4-Dimethylphenol 0.006    <
5 2013-04-23 2,4-Dimethylphenol 0.006    <
6 2013-05-22 2,4-Dimethylphenol 0.006    <
\end{CodeOutput}
\end{CodeChunk}
%
This data set identifies substances only by substance names. Values below the limit of quantification (LOQ) are indicated by a qualifier column.

(ii) \code{lc50}: This data consists of median acute lethal concentration for the water flea \textit{Daphnia magna} in 48 h tests ($LC_{50, D.magna, 48h}$) of 124 insecticides.
The data has been retrieved from the EPA ECOTOX database \citep{epa_2016}.
%
\begin{CodeChunk}
\begin{CodeInput}
R> data("lc50", package = "webchem")
R> head(lc50)
\end{CodeInput}
\begin{CodeOutput}
       cas        value
4  50-29-3    12.415277
12 52-68-6     1.282980
15 55-38-9    12.168138
18 56-23-5 35000.000000
21 56-38-2     1.539119
36 57-74-9    98.400000
\end{CodeOutput}
\end{CodeChunk}
%
This data set identifies the substances only by CAS numbers.

@@

One possibility to resolve this, is to use different chemical
identifiers allowing easy identification.  There are several
identifiers available, e.g., registry numbers like CAS or EC, database
identifiers like PubChemCID \citep{Kim_2016} or ChemSpiderID
\citep{pence_chemspider:_2010}, line notations like SMILES
\citep{Weininger_1990}, InChI and InChIKey
\citep{Heller_McNaught_Pletnev_Stein_Tchekhovskoi_2015}.  In this
first example we query several identifiers to create a table that can
be used as (i) supplemental information to a research article or (ii)
to facilitate subsequent matching with other data.

@@

First we search for ETOX internal ID numbers using the substance names:
%
\begin{CodeChunk}
\begin{CodeInput}
R> subs <- unique(jagst$substance)
R> ids <- get_etoxid(subs, match = "best")
R> head(ids)
\end{CodeInput}
\begin{CodeOutput}
  etoxid                           match distance                  query
1   8668     2,4-Dimethylphenol ( 8668 )        0     2,4-Dimethylphenol
2   8494 4-Chlor-2-methylphenol ( 8494 )        0 4-Chlor-2-methylphenol
3   <NA>                            <NA>     <NA>     4-para-nonylphenol
4   8397                Atrazin ( 8397 )        0                Atrazin
5   7240                 Benzol ( 7240 )        0                 Benzol
6   7331        Desethylatrazin ( 7331 )        0        Desethylatrazin
\end{CodeOutput}
\end{CodeChunk}
%
Only three substances could not be found in ETOX.  Here we specify
that only the \code{"best"} match (in terms of the Levenshtein
distance between query and results) is returned.  A manual check
confirms appropriate matches.  Other options include: \code{"all"} --
returns all matches; \code{"first"} -- returns only the first match
(not necessarily the best match); \code{"ask"} -- this enters an
interactive mode, where the user is asked for a choice if multiple
matches are found and \code{"na"} which returns \code{NA} in case of
multiple matches.

We use these data to retrieve basic information on the substances.
%
\begin{CodeChunk}
\begin{CodeInput}
R> etox_data <- etox_basic(ids$etoxid)
\end{CodeInput}
\end{CodeChunk}
%
\pkg{webchem} always returns a named list (one entry for each substance) and the available information content can be very voluminous.
Therefore, we provide extractor functions for the common identifiers: CAS, SMILES and InChIKeys.
%
\begin{CodeChunk}
\begin{CodeInput}
R> etox_cas <- cas(etox_data)
R> head(etox_cas)
\end{CodeInput}
\begin{CodeOutput}
       8668        8494        <NA>        8397        7240        7331 
 "105-67-9" "1570-64-5"          NA "1912-24-9"   "71-43-2" "6190-65-4" 
\end{CodeOutput}
\end{CodeChunk}
%
A variety of data are available and we cannot provide extractor
functions for each of those.  Therefore, if users need to extract
other data, they have to write simple extractor functions (see the
following examples).

In the same manner, we can now query other identifiers from another
source using these CAS numbers (see Figure~\ref{fig:fig1}), like
PubChem
%
\begin{CodeChunk}
\begin{CodeInput}
R> cids <- get_cid(etox_cas)
R> pc_data <- pc_prop(cids, properties = "CanonicalSMILES")
R> pc_smiles <- smiles(pc_data)
\end{CodeInput}
\end{CodeChunk}
%
or ChemSpider
%
\begin{CodeChunk}
\begin{CodeInput}
R> csids <- get_csid(etox_cas, token = token)
R> cs_data <- cs_compinfo(csids, token = token)
R> cs_inchikey <- inchikey(cs_data)
\end{CodeInput}
\end{CodeChunk}
%
Finally, we combine the queried data into one data frame
%
\begin{CodeChunk}
\begin{CodeInput}
R> res <- data.frame(name = subs, cas = etox_cas, smiles = pc_smiles, 
+    cid = pc_data$CID, inchikey = cs_inchikey, csid = cs_data$csid, 
+    stringsAsFactors = FALSE)
\end{CodeInput}
\end{CodeChunk}
%
Note that in order to use the ChemSpider functions, a personal
authentication key (\code{token}) is needed, which can be retrieved
from the ChemSpider web page.  Finally, we obtain a compound table
containing many different identifiers (Table~\ref{tab:comptable}),
allowing easy identification and merging with other data sets, e.g.,
the \code{lc50} data set based on CAS.

\begin{table}[t!]
\centering
\begin{tabular}{llllll}
  \hline
Name & CAS & SMILES & CID & InChIKey & CSID \\ 
  \hline
2,4-Dimethylphenol & 105-67-9 & CC1=CC(... & 7771 & KUFFULV... & 13839123 \\ 
  4-Chlor-2-methylphenol & 1570-64-5 & CC1=C(C... & 14855 & RHPUJHQ... & 14165 \\ 
  4-para-nonylphenol & -- & -- & -- & -- & -- \\ 
  Atrazin & 1912-24-9 & CCNC1=N... & 2256 & MXWJVTO... & 2169 \\ 
  Benzol & 71-43-2 & C1=CC=C... & 241 & UHOVQNZ... & 236 \\ 
  Desethylatrazin & 6190-65-4 & CC(C)NC... & 22563 & DFWFIQK... & 21157 \\ 
   \hline
\end{tabular}
\caption{Identifiers for the \code{jagst} data sets as queried with
  \pkg{webchem}. Only the first 6 entries are shown. For SMILES and
  InChIKey only the first 7 characters are shown. -- = not found.}
\label{tab:comptable}
\end{table}

\subsection[Toxicity of different pesticide groups]{Toxicity of different pesticide groups}
Another question we might ask is \emph{How does toxicity vary between
  insecticide groups?}  Answering this question would require tedious
lookup of insecticide groups for each of the 124 CAS numbers in the
\code{lc50} data set.  The Compendium of Pesticide Common Names
\citep{wood} contains such information and can be easily queried using
CAS numbers with \pkg{webchem}:
%
\begin{CodeChunk}
\begin{CodeInput}
R> aw_data <- aw_query(lc50$cas, type = "cas")
\end{CodeInput}
\end{CodeChunk}
%
To extract the chemical group from the retrieved data set, we write a simple extractor function and apply this to the retrieved data:
%
\begin{CodeChunk}
\begin{CodeInput}
R> igroup <- sapply(aw_data, function(y) y$subactivity[1])
R> igroup[1:3]
\end{CodeInput}
\begin{CodeOutput}
                                  50-29-3 
            "organochlorine insecticides" 
                                  52-68-6 
               "phosphonate insecticides" 
                                  55-38-9 
"phenyl organothiophosphate insecticides" 
\end{CodeOutput}
\end{CodeChunk}
%
Figure~\ref{fig:fig2} displays the result after additional data
cleaning (see the supplementary material for the full code).  Overall, it
took only 5 \proglang{R} statements to retrieve, clean and plot the
data using \pkg{ggplot2} \citep{ggplot2}.

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.7\textwidth, trim = 0 10 0 5, clip]{plot_lc50-1} 
\caption{Toxicity of different pesticide groups. LC\textsubscript{50}
  values have been retrieved from the EPA ECOTOX database, chemical groups
  from the Compendium of Pesticide Common Names \citep{wood}.}
\label{fig:fig2}
\end{figure}
